---
title: "'More than One Race' and Erasure in Demographics"
format: html
bibliography: references.bib
---

# Background

```{r setup, include=FALSE}
#Set Up Environment
source("~/Documents/GitHub/Portfolio/Portfolio/scripts/run_instructions.R")
#Run Transformations
source("~/Documents/GitHub/Portfolio/Portfolio/scripts/MultiRacial_Exploratory_Data_Analysis.R")

```

## Seed of the Question:

Stemmed from early work on self-identification and census data; conversations with colleagues and collaborators.

## Question:

How does the erasure of "racialized" experiences in America through the function of collecting and reducing race to "only" or "more than one" perpetuates harm (such as under counting or devaluing)

## Literature Review:

Dr. Camara Jones (2000) "Levels of Racism: A Theoretic Framework and a Gardener's Tale"@levelso2000:

-   Institutional Racism
-   Mediated Racism
-   Internalized Racism

Dr. Kenneth Prewitt (2018):"The Census Race Classification: Is it Doing Its Job?"@prewitt2018

-   "racialized" immigrant groups and "whitened" immigrant groups
-   health, education, and employment
    -   Note: often the demographer's key metrics to show cohort effects
-   strong racial self-identification among individuals who identify as white, Black, or Asian; lower racial self-identification among individuals who identify as Hispanic/Latino/a/x/e, Indigenous, and "Biracial"
    -   Note: different enough for lowering reliability of race data.

        -   Question: Could this be a design feature of enumeration efforts and trauma loop?
-   Quick list of articles regarding "racialized" groups and classification for future exploration, thoughts synthesizing:
    -   Drs. Patino, Gómez & Pinckney (2025): "Historical and Sociolinguistic Insights for Recreation Researchers: Understanding the Terms and Evolution of Hispanic, Latino, and Latinox"@patino2025
        -   Language and self-identification; Labeling
    -   Dr. Ivis García (2020): "Cultural Insights for Planners: Understanding the Terms Hispanic, Latino, and Latinx"@garcía2020
        -   Policy implications and self-identification
    -   Dr. Livny (2026) "Don’t count me out: erasure of ethnicity and ethnic groups from national censuses"@livny2026
        -   Systematic Inclusion and Representation
    -   Drs on behalf of Society of Epidemiologic Research Diversity and Inclusion Committee "Assessing Representation and Perceived Inclusion Among Members of the Society for Epidemiologic Research"@devilbiss2020
        -   Political Affiliation, Self-Identification, and Measured Social Participation

### Wire-Framing Framework:

\~\~\~SYSTEMS THINKING; language and bias

| Keyword           | Scholar(s)       | Constructs                |
|-------------------|------------------|---------------------------|
| "Governmentality" | Michel Foucault  | knowledge, power, systems |
|                   |                  |                           |
|                   |                  |                           |
|                   |                  |                           |

```{r Table, echo=FALSE}

#Create Table for ONLY Version
king_census_trace_race  %>%
  mutate(
    # IMPORTANT: Re-Label Racial Identities
    variable = case_when(
      variable == "AAA" ~ "Asian or Asian American",
      variable == "AIANI" ~ "American Indian, Alaskan Native, or Indigenous",
      variable == "BAAA" ~ "Black, African, or African American",
      variable == "HL" ~ "Hispanic or Latino/a/x/e",
      #MENA NOT PROJECTED OR ESTIMATED  
      variable == "NHPI" ~ "Native Hawaiian or Pacific Islander",
      variable == "W" ~ "White",
      variable == "SORA" ~ "Some Other Race Alone",
      variable == "TOMR" ~ "Two or More Races",      
      TRUE ~ variable
    ),
    # Estimate and MOE
    estimate = round(estimate, 2),
    moe = round(moe, 2),
    # Cleaner Est & MOE
    est_moe = paste0(
      round(estimate, 1),
      " (±",
      round(moe, 1),
      ")"
    )
  ) %>%
  select(GEOID, NAME, variable, est_moe) %>%
  datatable(
    filter = "top",   # adds filter boxes
    options = list(
      pageLength = 20,
      autoWidth = TRUE
    ) 
    #%>%
    #   select(GEOID, NAME, variable, est_moe) %>%
    #   reactable(
    #     filterable = TRUE,
    #     searchable = TRUE
    # )
  )


```

```{r Map, echo=FALSE}

ggplot(wa_race_zip) +
  geom_sf(aes(fill = estimate)) +
  theme_minimal()

```
