---
title: "Prime_Lab_SkillsAssessment"
format: html
author: "Cathea M. Carey, MPH"
date: "12/12/25"
output:
  pdf_document:
    number_sections: true
    toc: true
bibliography: references.bib
---

# Background

Behavioral Health (BH) is defined as X. Commonly delivered in Y settings, and in the United States' health care system often are inaccessible to populations who would benefit from BH interventions due to reason A- (cite), reason B (cite), and reason C (cite).

Primary Care (PC) is often a seen as a secondary health care setting that provides primary intervention for both physical and behavioral needs. PC access and utilization are associated with better outcome A - C.

BH interventions in PC have the shown potential to 1) loneliness and 2) depression as this setting have the added benefit of primary intervention or initial identification while the patient is potentially more motivated in PC rather than waiting for a referral (cite) or co-located BH/MH on-site (cite).

This small randomized control trial (RCT) aims to understand the feasibility of a BH intervention in Primary Care. The sample consists of 40 participants, equally weighted in the intervention (BH) and control (Treatment-as-usual or TAU). The data-set consists of intervention arm (BH/TAU), participant demographics (Age, Race/Ethnicity, Gender, and Income), measurement of outcomes (primary = loneliness; secondary = depression) at 3 time points (baseline, 3 months follow-up, and 6 months follow-up), and attendance for all 6 intervention session (Intervention arm = BH only).

# Decision Logs

## Pre-Extract, Transform, and Load (ETL) - Prior to Establishing the Environment

The Pre-Work before starting the assessment began with reviewing the "Skills_Assessment_Instructions.docx" and "data_dictionary.xlsx" files. During Pre-Work for the Assessment, the following thoughts came to mind:

-   What are the standard bins or categories for the following known continuous variables to measure the outcomes (primary = loneliness; secondary = depression):

    -   Age (for this research aim: Behavioral Health Interventions in Primary Care),
    -   PHQ-9 (measure for secondary outcome = depression),
    -   and UCLA - L20 (measure for primary outcome = loneliness)

-   **NOTE:** Gender outputs - Question what source(s) for outputs or standards for Gender were used? I've seen Gender reported as Gender Identify (a specific operationally defined construct with consistent industry standards) and Sexual Identity. I would suggest aligned with Office of Management and Budget's demographic standards back in 2024.

-   Additional Calculated field for Race/Ethnicity since a "self-described" output would likely result in small Ns.

-   **Caution:** Given the small sample size, external publications of findings should follow data suppression standards (such as Journals or Governmental entities).

```{r ETL Part 1, echo=FALSE, warning=FALSE, include=FALSE}
## Data Extraction, Transformation, and final Load for Analysis

# Packages
          library(readxl)
          library(purrr)
          library(dplyr)
          library(tidyr)
          library(janitor)
          library(gt)
          library(gridExtra)
          library(quantreg)
          library(tibble)
          library(broom)
          library(effectsize)
          library(ggplot2)

# Load files and establish environment (optional)

# NOTE - Data is on external drive

# Check driver for files

list.files("/Volumes/LaCie/DropBox/Github/Skills_Assessment_Dec_2025")

# Check xlsx for sheet names
file_path <- "/Volumes/LaCie/DropBox/Github/Skills_Assessment_Dec_2025/pilot_rct_data.xlsx"  

sheet_names <- excel_sheets(path = file_path)

# Add sheets as list then transform into DF (left join)

datalist <- lapply(sheet_names, function(sheet_name) {
  read_excel(path = file_path, sheet = sheet_name)
})

# Intermediate Data Frames for initial cleaning and final analysis

df_outcomes <- as.data.frame(datalist[[1]])

df_attendance <- as.data.frame(datalist[[2]])

# Check Heads
head(df_attendance)

head(df_outcomes)

# Final Data Frame for final analysis featuring left join of df_attendance to df_outcomes

df <- left_join (df_outcomes, df_attendance, by = "record_id")

# Check Head

head(df) 
### END OF ETL PART 1
```

## Post-ETL

### Make Calculated Fields

-   Age (bins)
-   Race/Ethnicity (New Variable Names/Labels)
    1.  Preserves the Categories with new field; one-hot encoding for select all option not needed
-   PHQ-9 Total (by time points)
-   PHQ-9 X Sub-categories (by time points)
    -   [PHQ-9: Validity of a Brief Depression Severity Measure](https://onlinelibrary.wiley.com/doi/full/10.1046/j.1525-1497.2001.016009606.x) (@kroenke2001)
        -   Validation suggests bins of PHQ-9 scores of 5== mild, 10 == moderate, 15 ==moderately severe, and 20 severe depression
-   UCLA-LS-20 Total (by time points)
-   UCLA-LS-20 X Sub-categories (by time points)
    -   [UCLA - LS Scale Article](https://www.cambridge.org/core/journals/psychological-medicine/article/measuring-loneliness-a-headtohead-psychometric-comparison-of-the-3-and-20item-ucla-loneliness-scales/CF99BEB620AF61EAEDCF7B360312F994) (@gosling2024)
        -   Sensitivity analysis suggests exploring cutoff at \>= 39; \>=43; and \>=53
-   Attendance (bins)
    1.  Preserve time-point attendance; additional overall attendance output for high-level trend (signal and keeps option for treatment dosage effects).
        1.  Prop = (full dose/'yes'+ 1/2 partial dose/'partial')/ total dosage (all 6 sessions)

Note: Create a Flag for records missing 1 or more elements (UCLA-LS-20; PHQ-9; Race/Ethnicity)

```{r ETL Part 2, echo=FALSE, warning=FALSE, results='hide'}
# Create Calculated fields (Age, Race/Ethnicity, PHQ-9, UCLA - LS- 20, and creating Dosages as a function of Attendance in Invention Tx arm only)

# Age (bins) - From Continuous Variable to Bins; 

hist(df$age) #review original field

# New Field
df$age_range <-  cut(df$age, # Suggested based on Histogram == 20 - 34, 35 - 49, 50 - 64, 64 and up
                     breaks=c(20, 35, 49, 64, 80),
                     labels=c('20 - 34 years old', '35 - 49 years old', '50 - 64 years old', '64 years old or older'))

table(df$age_range) #spot check

table(df$age) #reference line for above spot check, note best practice would be to create a separate subset or frame with Primary Key (PK), age, and age_range

#Race/Ethnicity (New Variable Names/Labels)

table(df$race_ethnicity) #review original field

table(df$race_nl_specify) #review original field

# New Field
df$raceethnicity_recoded <- case_when(
  df$race_ethnicity == "southeast Asian" ~ "Asian American", #fast hard coding of n = 2 fields to save time, note best practice is to still review in the same fashion as age
  df$race_ethnicity == "Nigerian" ~ "Black/AA",
  TRUE ~ df$race_ethnicity
)
table(df$raceethnicity_recoded)


# PHQ-9 Total (by time points)

# Find column numbers and quick calculate

which(names(df) %in% c("ucla_b_1", "ucla_3_1", "ucla_6_1","phq_b_1", "phq_3_1", "phq_6_1", "attend_s1", "attend_s6")) #Use this hack to file the column numbers to reference below

df$phq_b_total <- rowSums((df[, 68:76])) # Baseline PHQ-9 total

hist(df$phq_b_total) #review original field

df$phq_3_total <- rowSums((df[, 77:85])) # 3 month PHQ-9 total

hist(df$phq_3_total) #review original field

df$phq_6_total <- rowSums((df[, 86:94])) # 6 month PHQ-9 total

hist(df$phq_6_total) #review original field

# PHQ-9 X Sub-categories (by time points) PHQ-9 scores of 5== mild, 10 == moderate, 15 ==moderately severe, and 20 severe depression
df <- df %>% # Baseline PHQ-9 total
  dplyr::mutate(
    phq_b_cat = case_when(
      phq_b_total <= 5 ~ "Mild Depression",
      phq_b_total <= 10 ~ "Moderate Depression",
      phq_b_total <= 15 ~ "Moderate Severe Depression",
      phq_b_total <= 20 ~ "Severe Depression",  
      TRUE ~ NA_character_
    )
  )
df <- df %>% #Factor levels
  mutate(
    phq_b_cat = factor(
      phq_b_cat,
      levels = c("Mild Depression", "Moderate Depression","Moderate Severe Depression","Severe Depression"),
      ordered = TRUE
    )
  )

df <- df %>% # 3 month PHQ-9 total
  dplyr::mutate(
    phq_3_cat = case_when(
      phq_3_total <= 5 ~ "Mild Depression",
      phq_3_total <= 10 ~ "Moderate Depression",
      phq_3_total <= 15 ~ "Moderate Severe Depression",
      phq_3_total <= 20 ~ "Severe Depression",  
      TRUE ~ NA_character_
    )
  )
df <- df %>% #Factor levels
  mutate(
    phq_3_cat = factor(
      phq_3_cat,
      levels = c("Mild Depression", "Moderate Depression","Moderate Severe Depression","Severe Depression"),
      ordered = TRUE
    )
  )

#df$phq_6_cat <- rowSums((df[, 86:94])) 

df <- df %>% # 6 month PHQ-9 total
  dplyr::mutate(
    phq_6_cat = case_when(
      phq_6_total <= 5 ~ "Mild Depression",
      phq_6_total <= 10 ~ "Moderate Depression",
      phq_6_total <= 15 ~ "Moderate Severe Depression",
      phq_6_total <= 20 ~ "Severe Depression",  
      TRUE ~ NA_character_
    )
  )
df <- df %>% #Factor levels
  mutate(
    phq_6_cat = factor(
      phq_6_cat,
      levels = c("Mild Depression", "Moderate Depression","Moderate Severe Depression","Severe Depression"),
      ordered = TRUE
    )
  )

# UCLA-LS-20 Total (by time points) 
df$ucla_b_total <- rowSums((df[, 8:27])) # Baseline UCLA-LS-20 total

hist(df$ucla_b_total) #review original field

df$ucla_3_total <- rowSums((df[, 28:47])) # 3 month UCLA-LS-20 total

hist(df$ucla_3_total) #review original field

df$ucla_6_total <- rowSums((df[, 48:67])) # 6 month UCLA-LS-20 total

hist(df$ucla_6_total) #review original field


# UCLA-LS-20 X Sub-categories (by time points) cutoff at >= 39; >=43; and >=53
df <- df %>% # >= 39; >=43; and >=53
  mutate(  # Baseline UCLA-LS-20 
    uclas_b_cut39 = ifelse(ucla_b_total >= 39, 1, 0),
    uclas_b_cut43 = ifelse(ucla_b_total >= 43, 1, 0),
    uclas_b_cut53 = ifelse(ucla_b_total >= 53, 1, 0)
  )

df <- df %>% # >= 39; >=43; and >=53
  mutate(  # 3 month UCLA-LS-20 
    uclas_3_cut39 = ifelse(ucla_3_total >= 39, 1, 0),
    uclas_3_cut43 = ifelse(ucla_3_total >= 43, 1, 0),
    uclas_3_cut53 = ifelse(ucla_3_total >= 53, 1, 0)
  )

df <- df %>% # >= 39; >=43; and >=53
  mutate(  # 6 month UCLA-LS-20
    uclas_6_cut39 = ifelse(ucla_6_total >= 39, 1, 0),
    uclas_6_cut43 = ifelse(ucla_6_total >= 43, 1, 0),
    uclas_6_cut53 = ifelse(ucla_6_total >= 53, 1, 0)
  )

# Attendance (bins) - prepare for factor analysis of "treatment dosage"

# Test look into Attendance as a continuous variable by re-coding qualitative flags into quantitative to treat dosage as a product of attendance assumed linear

values <- c("no" = 0, "partial" = 0.5, "yes" = 1)

# Dosage Total

df$trt_dosage_total <- rowSums(
  sapply(df[, 95:100], function(x) values[x]),
  na.rm = TRUE
)

hist(df$trt_dosage_total) #review original field

# Dosage Category (labeled) - Sample label that should be vetted

bh_dosage_total <- ncol(df[, 95:100])

df$dosage_prop <- (rowSums(df[, 95:100] == "yes", na.rm = TRUE) + #treats yes as full dosage, partial as half dosage, no as missed dosage
                     0.5 * rowSums(df[, 95:100] == "partial", na.rm = TRUE)) / bh_dosage_total

df$trt_dosage <- cut(
  df$dosage_prop,
  breaks = c(-0.01, 0, 0.33, 0.66, 1),   # include 0 as its own interval
  labels = c("Control", "Minimal Treatment", "Partial Treatment", "Full Treatment")
)

# Make subsets for potential comparison analysis and data quality check

subset_BHarm_df <- subset(df, arm == "Intervention") #spot check to make sure left join was successful = correct attendance for intervention only

subset_TAUarm_df <- subset(df, arm == "Control") #spot check to make sure left join was successful = correct attendance for intervention only


### END OF ETL PART 2
```

### Tables 1: Pilot Sample Characteristics (N=40)

```{r Table 1, echo=FALSE, warning=FALSE}
# Suggested: Make a Table 1 (version is for internal but include total and use data suppression for small Ns)

#Overall Dosage for Summary and to answer Q1-3

hist(subset_BHarm_df$trt_dosage_total) #review original field and print
#average
avg_trt_dosage <- mean(subset_BHarm_df$trt_dosage_total)
#standard deviation
sd_trt_dosage <- sd(subset_BHarm_df$trt_dosage_total)
#median
mdn_trt_dosage <- median(subset_BHarm_df$trt_dosage_total)

# Now at the record level for Df summary
subset_BHarm_df$all_dosage <- subset_BHarm_df$dosage_prop == 1 #count full dosage
full_dosage_rate <- mean(subset_BHarm_df$all_dosage) # full dosage rate
full_dosage_n <- sum(subset_BHarm_df$all_dosage) # sum of full dosage for stat
full_dosage_n_total <- nrow(subset_BHarm_df) # another way to count denominator for retention


# Table 1 Var
demo_char_table1 <- df %>%
  select(arm, age_range, gender, raceethnicity_recoded, phq_b_total, phq_3_total, phq_6_total, ucla_b_total, ucla_3_total, ucla_6_total,trt_dosage, trt_dosage_total, income)

# Level with intervention arm
demo_char_table1$arm <- factor(demo_char_table1$arm, levels = c("Intervention", "Control"))

# Prep diff var types
categorical_vars <- c("age_range", "gender", "raceethnicity_recoded", "trt_dosage")
continuous_vars <- c("phq_b_total", "phq_3_total","phq_6_total","ucla_b_total", "ucla_3_total","ucla_6_total","income", "trt_dosage_total")

# Cat variables as N (%)
cat_table <- demo_char_table1 %>%
  select(arm, all_of(categorical_vars)) %>%
  pivot_longer(cols = -arm, names_to = "characteristic", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(characteristic = recode(characteristic,
                                 age_range = "Age (bin)",
                                 gender = "Gender",
                                 raceethnicity_recoded = "Race/Ethnicity",
                                 trt_dosage = "Intervention Dosage")) %>%
  group_by(characteristic, value, arm) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(characteristic, arm) %>%
  mutate(total = sum(n),
         n_perc = paste0(n, " (", round(100*n/total,1), "%)")) %>%
  select(characteristic, value, arm, n_perc) %>%
  pivot_wider(names_from = arm, values_from = n_perc, values_fill = "0 (0%)") %>%
  arrange(characteristic)

# Cont. variables as Mean (SD)
cont_table <- demo_char_table1 %>%
  select(arm, all_of(continuous_vars)) %>%
  pivot_longer(cols = -arm, names_to = "characteristic", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(characteristic = recode(characteristic,
                                 phq_b_total = "PHQ-9 at baseline (Mean(SD))",
                                 phq_3_total = "PHQ-9 at 3 months (Mean(SD))",
                                 phq_6_total = "PHQ-9 at 6 months (Mean(SD))",                                 
                                 ucla_b_total = "UCLA - LS - 20 at baseline (Mean(SD))",
                                 ucla_3_total = "UCLA - LS - 20 at 3 months(Mean(SD))",
                                 ucla_6_total = "UCLA - LS - 20 at 6 months (Mean(SD))",
                                 income = "Annual Household Income in USD (Mean(SD))",
                                 trt_dosage_total = "Total Doseage of Treatment (Mean(SD))"
  )) %>%
  group_by(characteristic, arm) %>%
  summarise(mean_sd = paste0(round(mean(value),1), " (", round(sd(value),1), ")"),
            .groups = "drop") %>%
  pivot_wider(names_from = arm, values_from = mean_sd)

# Combine both type of vars tables
draft_demo_table <- bind_rows(cat_table, cont_table)

# Clean column names
draft_demo_table <- draft_demo_table %>% janitor::clean_names()

# Final Table
draft_demo_table %>%
  gt() %>%
  tab_header(title = "Demographic and Baseline Characteristics by Treatment") %>%
  cols_label(
    value = "Characteristics",
    intervention = "Intervention (N(%))",
    control = "Treatment-as-Usual (N(%))"
  ) %>%
  fmt_missing(columns = everything(), missing_text = "")

# Ungroup if it's grouped — grouped tables sometimes don't print well
#df <- draft_demo_table %>% ungroup()

png("table_1_output.png", width = 1000, height = 800)

grid.table(df)

dev.off()
##END
```

# Skills Assessment Documentation:

## Rationale for Specific Output of R

Below is the R Code created to assess the following skills of the author (myself):

-   Ability to handle and prepare real-world data,

-   strategies behind analysis choices,

-   interpretation of results in context of a small, pilot RCT,

-   and overall clarity and accessibility of the communication of the above.

I believe this documentation shows through the RMD function 1) in line comments and choices for project/team documentation and replicability, 2) use of a repository to decrease staff time needed to tailor communication findings to various outputs, and 3) RMD output with PowerPoint Presentation (PPT), Word Document or PDF (industry standard or open source static documents to review), and Webpage (HMTL) for easy intra-team QA/QI for the shared potential increase audience comprehension through multiple outputs with click of a button.

## **Key Research Questions**

1.  How well were participants retained? What is implied about the feasibility of a larger trial?

    A:Among n=20 participants, treatment dosage average treatment dosage was 4.15 (SD = 0.919) out of 6 sessions. *This suggests across the entire treatment dosage (6 sessions) attendance and presumed dosage of treatment is feasible in for this intervention design.* Qualitatively, n = 13 (65%) completed roughly 4 or more sessions (Full Treatment).

    Note: Sample Characteristics of the N= 40 are proportions for strata across intervention arm except for Age, Gender, Race/Ethnicity, and average annual household income Income (USD). Average scores for PHQ-9 and UCLA-LS-20 were similar across intervention arm, both at moderate severe depression and between 43 and 52 loneliness score.

    **Test the effect of possibly differences in demographic characteristics (i.e., income) and their relationship with other covariates (Race/Ethnicity and Age) on the causal pathway.**

2.  Does the intervention seem to reduce loneliness (primary) or depression (secondary)?

    A: Cohen's d showed a small effect, further testing is recommended.

    Generalized Linear Models or Mixed Model Repeat Measure ([MMRM](https://openpharma.github.io/mmrm/latest-tag/)) or ANOVA to look into the variance for further exploration. Would suggest just starting with tailor test for small sample sizes - Wilcoxon Signed Rank Test and treat Outcome as continuous for an initial test for significance. Larger sample would warrant a GLM or model with less rigid parameter requirements.

3.  Are there participant characteristics (e.g., baseline loneliness or depression severity, attendance, or demographic characteristics) that appear to be associated with outcomes or that may influence how we interpret the intervention’s effects?

    Mixed Model Repeat Measure ([MMRM](https://openpharma.github.io/mmrm/latest-tag/)) with Demographic Characteristics Age, Race/Ethnicity, Gender, and Income since those may be the most different between arms (Fisher's Test?). Again - would want to refer to the assumed relationships and interactions (DAG == casual pathway).

```{r Test, echo=FALSE, warning=FALSE, include=FALSE}

## Code for Key Research Questions
# Question 1 (part A) How well were participants retained? 
#hist(subset_BHarm_df$attend_s1)
#hist(subset_BHarm_df$attend_s2)
#hist(subset_BHarm_df$attend_s3)
#hist(subset_BHarm_df$attend_s4)
#hist(subset_BHarm_df$attend_s5)
#hist(subset_BHarm_df$attend_s6)
# Question 1 (part B) What is implied about the feasibility of a larger trial?
avg_trt_dosage
sd_trt_dosage

# Question 2 Does the intervention seem to reduce loneliness (primary) or depression (secondary)?

# Cohen's d for effect size
cohens_d(ucla_6_total ~ arm, data = df)



# Paired Wilcoxon
#test0 <- wilcox.test(df$ucla_b_total, df$ucla_6_total, paired = TRUE, conf.int = TRUE, conf.level = 0.95) # Not completed


# Question 3 Are there participant characteristics (e.g., baseline loneliness or depression severity, attendance, or demographic characteristics) that appear to be associated with outcomes or that may influence how we interpret the intervention’s effects?
hist(df$income)

 income_plot <- ggplot(df, aes(x = income, fill = arm, color = arm)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(title = "Income by Arm", x = "Income", y = "Frequency") +
  theme_classic()
 
 ### END due to time CCAREY 12/12/25


# Paired Wilcoxon
test1 <- wilcox.test(df$ucla_b_total, df$ucla_6_total, paired = TRUE, conf.int = TRUE, conf.level = 0.95)


# Question 3 Are there participant characteristics (e.g., baseline loneliness or depression severity, attendance, or demographic characteristics) that appear to be associated with outcomes or that may influence how we interpret the intervention’s effects?

#rq(outcome ~ arm + age + gender + income, data = df, tau = 0.5)



```
